```
title: "Affymetrix Microarray Minimal pipeline"
author: "Simon Tomlinson 10/02/2021"
output: html_document
```
# Affymetrix Microarray Analysis Basic (Skeleton) Workflow 
##Load the required libraries & load the files for the workflow
```{r libload, echo=TRUE}
library(limma)
library(affy)
library(annotate)
library(mouse4302.db)# load chip-specific annotation
#install.packages("scatterplot3d",repo="http://cran.ma.imperia#l.ac.uk")
#Then load the library
library(scatterplot3d)
```

## Load the main data- commented code is just for information
```{r dataload, echo=TRUE}
# Load the target file into an AnnotatedDataFrame object
adf<-read.AnnotatedDataFrame("targets2.txt",header=TRUE,row.names=1,as.is=TRUE)
targets2<-read.table("targets2.txt",header=TRUE,as.is=TRUE)
row.names(targets2)<-targets2$Name
# Load the expression values of all the CEL files in the targets file
mydata <- ReadAffy(filenames = targets2$Filename)
# View a summary of the example data
sampleNames(mydata)
#Note that the sample order has changed to match the order specified in #targets2$Filename.
#The data is no longer alphabetical sorted by Filename!
# View a summary of the example data
mydata
```
## Build Quality Control Plots
```{r qc_plots1, echo=FALSE}
# Quality control plots
png("Bhist_before.png")
hist(mydata)
dev.off()

library(simpleaffy)
#data is an AffyBatch eg generated using ReadAffy
X.mas<-call.exprs(mydata,"mas5")
qcs<-qc(mydata,X.mas)
png("Bqc.png")
plot(qcs)
dev.off()

# And a boxplot with different colour per sample group
png("Bbox_before.png")
boxplot(mydata, col=targets2$Colour, las=2,names=targets2$Name,ylab="Log intensity",main="boxplot_before")
dev.off()
```
## Normalise the data using RMA
```{r normalise, echo=FALSE}
# Normalise the data using RMA
eset <- rma(mydata)
eset
# To obtain a matrix of the expression values, use exprs() 
values <- exprs(eset)
# Boxplot to observe the results of normalisation
# Notice differences with the boxplot from the raw data
png("Bbox_after.png")
boxplot(values, col=targets2$Colour,las=2,names=targets2$Name,ylab="Log intensity",main="boxplot_after")
dev.off()
```

## Plot Normalised Data
```{r plot_normalised, echo=FALSE}
# MA plot of the samples 
png("Bmvapairs1norm_normalised.png")
mva.pairs(values)
dev.off()
# The same plot for the non-normalised raw data
# Note that the mva.pairs call below only plots a few of the  #samples – you may wish to plot them all but this is slow
png("Bmvapairs1unnorm_non-normalised.png")
mva.pairs(pm(mydata))
dev.off()

```
## Plot Heatmap
```{r heatmap_normalised, echo=FALSE}
# To facilitate interpretation, let’s replace the columns # # header,currently
# displaying the filename, to show the name of each sample 
# (if you have a targets file)
colnames(values) <- rownames(pData(adf))
# Performs hierarchical clustering with average linkage based on
# Pearson’s Correlation Coefficient
hc<-hclust(as.dist(1-cor(values, method="pearson")), method="average")
png("Bhclust1.png")
plot(hc)
dev.off()
```
## Perform PCA
```{r pca_normalised, echo=FALSE}
# Perform PCA
values <- values + rnorm(values,mean=0, sd=0.1)
pca <- prcomp(t(values), scale=T)
# Plot the PCA results
png("scatter3D.png")
s3d<-scatterplot3d(pca$x[,1:3], pch=19, color=rainbow(1))
s3d.coords <- s3d$xyz.convert(pca$x[,1:3])
text(s3d.coords$x, s3d.coords$y, labels = colnames(values),pos = 3,offset = 0.5)
dev.off()
```

## Perform fold filtering
```{r fold_filtering, echo=TRUE}
#obtaining a matrix of expression values
exprsvals <- exprs(eset)
write.csv(exprsvals,file="exprsvals.csv", quote=F)
#RMA outputs log2 data while MAS5 outputs linear data
#To convert from log…
exprsvals10 <-2^exprsvals
#check conversion
exprsvals[1:10,]
#converted
exprsvals10[1:10,]
```

```{r fold_filtering, echo=TRUE}
#More fold filtering
#check order of sample names
mysamples <- sampleNames(eset)
#display the list
mysamples
#it is useful to obtain a vector of ProbeIDs here
probesets <- probeNames(mydata)
#display the first 10 ProbeSets
probesets[1:10]
```

```{r fold_filtering, echo=TRUE}
#Build final fold table
#Calculate the means
#Note mean of the log is not the same as the log of the mean!!
control.mean <- apply(exprsvals10[,c("GSM4423390_NCOR1-LYN_-_1_.CEL", "GSM4423391_NCOR1-LYN_-_2_.CEL","GSM4423392_NCOR1-LYN_-_3_.CEL")],1,mean)
case.mean <- apply(exprsvals10[,c("GSM4423393_NCOR1-LYN_+_1_.CEL", "GSM4423394_NCOR1-LYN_+_2_.CEL","GSM4423395_NCOR1-LYN_+_3_.CEL")],1,mean)
#calculate some fold changes
control_case <- control.mean/case.mean
```

```{r fold_filtering, echo=TRUE}
#build a summary table to hold all the data
all.data= cbind(control.mean,case.mean,control_case)
#Descending order
new.all.data=all.data[order(all.data[,3],decreasing=T),]
new.all.data[1:10]
#check the column names
colnames(all.data)
#write the table of means as an output
write.table(all.data,file="group_means.txt", quote=F,sep="\t",col.names=NA)
write.csv(all.data,file="group_means.csv", quote=F)
```

## Beginning statistical analysis
```{r limma_stats, echo=TRUE}
#Check original sample order
sampleNames(eset)
#Rename the samples
sampleNames(eset) <-c("B-1","B-2","B-3","B+1","B+2","B+3")
#Check the samples have renamed
sampleNames(eset)
```

```{r building_annotation, echo=TRUE}
##Building annotation for differential gene identification
#establish annotation for MOE430v2
#which annotation do we need
#modified from #http://gettinggeneticsdone.blogspot.co.uk/2012/01/annotating-limma-#results-with-gene.html
eset@annotation
#packages in the annotation package
#load chip-specific annotation
ls("package:mouse4302.db")
#build an annotation table
ID <- featureNames(eset)
Symbol <- getSYMBOL(ID, "mouse4302.db")
Name <- as.character(lookUp(ID, "mouse4302.db", "GENENAME"))
tmp <- data.frame(ID=ID, Symbol=Symbol, Name=Name, stringsAsFactors=F)
tmp[tmp=="NA"] <- NA #fix padding with NA characters 
#assign as feature data of the current Eset
fData(eset) <- tmp
```

## Statistical analysis using Limma
```{r limma_statistical_analysis, echo=TRUE}
#Build the design matrix
design <- model.matrix(~-1+factor(c(1,1,1,2,2,2)))
colnames(design) <- c("control","case")
#Check it makes sense
sampleNames(eset)
#output the design matrix
design
```

```{r limma_statistical_analysis, echo=TRUE}
#This instructs Limma which comparisons to make
contrastmatrix <- makeContrasts(control-case,levels=design)
contrastmatrix

#issue these commands to fit the model
#and make the contrasts
fit <- lmFit(eset, design)
fit2 <- contrasts.fit(fit, contrastmatrix)
#this last part essentially moderates the t-statistic using 
#the borrowed variance approach described in class
fit2 <- eBayes(fit2)

#get the results
topTable(fit2,coef=1,adjust="fdr")
myresults <-topTable(fit2,coef=1, adjust="fdr", number=nrow(eset))
write.table(myresults,"myresults.txt")

#make a venn diagram
clas <- classifyTestsF(fit2)
vennDiagram(clas)
```

```{r limma_statistical_analysis, echo=TRUE}
#make volvano page
dim(myresults)
require(EnhancedVolcano)
#Make a picture
png("volcano.png")
#https://www.rdocumentation.org/packages/EnhancedVolcano/versions/1.11.3/topics/EnhancedVolcano
EnhancedVolcano(myresults,lab = rownames(myresults),x = "logFC",y = "P.Value")
dev.off
```
## Carry out Functional Enrichment analysis
```{r functional_enrichment, echo=TRUE}
#load("Mm.h.all.v7.1.entrez.rds")
#But load of this type of file is not supported by all 
#versions of R- so we use readRDS instead
Mm.H <- readRDS("/shared_files/MSigDB/Mm.h.all.v7.1.entrez.rds") 
#Check that you have the required objects
ls()
#Show the full contents of the annotation package
ls("package:mouse4302.db")
#Show the annotation keys in this database
keytypes(mouse4302.db) 
#
#http://genomicsclass.github.io/book/pages/mapping_features.ht
#ml
sampleNames(eset)
```
## Process annotation for functional enrichment
```{r process_annotation_for_enrichment, echo=TRUE}

#Here we select from the annotation a number of keys with the primary key being PROBEID
res <- select(mouse4302.db, keys = rownames(eset), columns = c("ENTREZID", "ENSEMBL","SYMBOL"), keytype="PROBEID")
#View the top of the table
head(res)
#find the index of each row of the expression set in the #annotation object res
idx <- match(rownames(eset), res$PROBEID)
#Use the index to set the phenotypic data in the ExpressionSet
fData(eset) <- res[idx, ]
head(fData(eset), 10)
#Find all rows that don’t have an EntrezID and remove then
eset_t<-eset[is.na(fData(eset)$ENTREZID)==0,]
```


## Functional Enrichment Analysis
```{r convert_indicex, echo=TRUE}

#convert to indexes
H.indices <- ids2indices(Mm.H,fData(eset_t)$ENTREZID)
#Pick the most suitable enrichment analysis tool to find #enrichment signatures in the data and run this tool So:-

#I just run mroast here as an example- justify the selection of this method!

#if you want to run mroast
#results <-mroast(eset_t,index=H.indices,design=design,contrast=contrastmatrix[,1],adjust.method = "BH")
#if you want to run camera
#results <-camera(eset_t,index=H.indices,design=design,contrast=contrastmatrix[,1],adjust.method = "BH")
#if you want to run romer
results <-romer(eset_t,index=H.indices,design=design,contrast=contrastmatrix[,1],adjust.method = "BH")
#View the results
results
#Use help for other parameters. Note we might decide to use #exactly the same model as our differential gene analysis for #the enrichment analysis- in this case we can extract it from #the fit
#sv <- squeezeVar(fit$sigma^2,df=fit$df.residual)

write.table(results,"enrichment.txt",sep="\t")
#You can then examine the results in “enrichment.txt”.  It is a text file.  It can be downloaded to view in a spreadsheet such as Excel.
```

## Session Information
```{r session_info, echo=TRUE}

sessionInfo()

```

